Empirical Analysis of Performance Bottlenecks in Graph Neural Network Training with GPUs

### 2020-12-18 NEUROC/sec2-review-of-graph-neural-network.tex
更新典型图神经网络介绍章节。

\begin{table}[h]
    \caption{Notations}
    \label{tab:notations}
    \centering
    \small
    \begin{tabular}{p{3em}lp{35em}}
        \toprule
       Category & Symbol & Meaning \\
        \midrule
        \multirow[c]{4}{3em}{Graph Structure}& $\mathcal{G}=(\mathcal{V}, \mathcal{E})$ & The simple undirected input graph with the vertex set $\mathcal{V}$ and the edge set $\mathcal{E}$. \\
        & $v_x$ & The $x$-th vertex of the input graph. \\
        & $e_{x,y}$ & The edge pointing from $v_x$ to $v_y$ of the input graph. \\
        & $\mathcal{N}(v_x)$ & The adjacency set of $v_x$ in the input graph. \\ \midrule
        \multirow[c]{6}{3em}{GNN Definition}& $L$ & The number of GNN layers. \\
        & $K$ & The number of heads in a GNN layer. \\
        & $\phi^l$ & The messaging function of the GNN layer $l$. \\
        & $\Sigma^l$ & The aggregation function of the GNN layer $l$. \\
        & $\gamma^l$ & The vertex updating function of the GNN layer $l$. \\ 
        & $\phi^{l,i}$/$\Sigma^{l,i}$/$\gamma^{l,i}$ & The messaging/aggregation/updating function of the $i$-th sub-layer of the GNN layer  $l$.\\
        & $\textcolor{blue}{\boldsymbol{W}^l, \boldsymbol{W}^{(k)}/\boldsymbol{b}, \boldsymbol{a}}$ & The matrices/vectors represented by the blue characters are the weight matrices/vectors that need to be learned in the GNN. \\  \midrule
        \multirow[c]{8}{3em}{Vectors}& $\boldsymbol{v}_x$ & The feature vector of the vertex $v_x$. \\
        & $\boldsymbol{e}_{x,y}$ & The feature vector of the edge $e_{x,y}$.  \\
        & $\boldsymbol{h}_x^{l}$ &  The {input} hidden vector of the graph neuron corresponding to $v_x$ in the GNN layer $l$. \\
        & $\boldsymbol{h}_x^{l+1}$ &  The {output} hidden vector of the graph neuron corresponding to $v_x$ in the GNN layer $l$.\\
        & $\boldsymbol{m}_{x,y}^l$ & The message vector of the edge $e_{x,y}$ outputted by $\phi^l$ of the GNN layer $l$. \\
        & $\boldsymbol{s}_{x}^l$ & The aggregated vector of the vertex $v_x$ outputted by $\Sigma^l$ of the GNN layer $l$. \\
        & $\boldsymbol{h}_{x}^{l,i}$/$\boldsymbol{m}_{x,y}^{l,i}$/$\boldsymbol{s}_{x}^{l,i}$ & The hidden/message/aggregated vector of the vertex $v_x$ outputted by $\gamma^{l,i}$/$\phi^{l,i}$/$\Sigma^{l,i}$ of the $i$-th sub-layer of the GNN layer $l$. \\
        & $d_{in}$, $d_{out}$ &  The dimensions of the input/output hidden vectors of a graph neuron. \\
        \bottomrule

@@

Graph neurons are building blocks of a GNN.
%
A GNN layer consists of $|\mathcal{V}|$ graph neurons.
%
Each vertex corresponds to a graph neuron.
%
A graph neuron is a small neural network.
%
For the graph neuron corresponding to $v_x$ at the layer $l$, it receives the hidden vector of itself $\boldsymbol{h}^l_x$ and the hidden vectors of its adjacent vertices $\boldsymbol{h}^l_y$ with $v_y \in \mathcal{N}(v_x)$ from the previous GNN layer\footnote{For the GNN layer 0, graph neurons receive input feature vectors, i.e., $\boldsymbol{h}^0_i=\boldsymbol{v}_i$}.
%
The graph neuron of $v_x$ aggregates the received hidden vectors, applies non-linear transformations, and outputs a new hidden vector $\boldsymbol{h}_x^{l+1}$.

@@

Different GNNs have different definitions of the three functions.
%
We regard $\phi$ and $\Sigma$ as \emph{edge calculation} functions, since they are conducted over every edge in $\mathcal{G}$.
%
We regard $\gamma$ as the \emph{vertex calculation} function, as it is conducted over every vertex in $\mathcal{G}$.
%
\tablename~\ref{tab:gnn_overview_edge} and \tablename~\ref{tab:gnn_overview_vertex} list the edge calculation functions and the vertex calculation functions of typical GNNs, respectively.

It is noteworthy that some complex GNNs like GAT, GaAN and ChebNet require more than one message passing phase in each GNN layer.
%
We regard every message passing phase in a GNN layer as a \emph{sub-layer}.
%
Take GAT \cite{huang2018_gat} as an example.
%
A GAT layer has two sub-layers and each sub-layer has $K$ heads under the multi-head attention mechanism.
%
For the GAT layer $l$, it first transforms every hidden vector received from the previous GAT layer $\boldsymbol{h}^l_x$ with a weight matrix $\textcolor{blue}{\boldsymbol{W}^{l}_{(k)}}$ for each head $k$ ($\parallel$ represents the concatenation operator):
%
\begin{equation}
   \hvec{x} = \parallel_{k=1}^K \Param{\MyMat{W}^l_{(k)}}\MyVec{h}^l_x, \quad \text{denoting } \hvec{x}[k]=\Param{\MyMat{W}^l_{(k)}}\MyVec{h}^l_x.
\end{equation}
%
The GAT layer $l$ then calculates the hidden vector $\MyVec{h}^{l,0}_x$ of every vertex $v_x$ in the first sub-layer as:
%
\newcommand{\GATCalcWeight}{\exp(LeakyReLU(\Param{\MyVec{a}}^T[\hvec{y}[k] \parallel \hvec{x}[k]]))}
\begin{equation}
    \begin{aligned}
        \MyVec{m}^{l,0}_{y,x} & = \phi^{l,0}(\MyVec{h}^l_y, \MyVec{h}^l_x, \MyVec{e}_{y,x}, \hvec{y}, \hvec{x}) = \parallel_{k=1}^{K}\GATCalcWeight, \\
        \MyVec{s}^{l,0}_{x} &= \mathlarger{\Sigma}_{v_y \in \mathcal{N}(v_x)}{\MyVec{m}^{l,0}_{y,x}}, \\
        \MyVec{h}^{l,0}_{x} &= \gamma^{l,0}(\MyVec{h}^l_x, \MyVec{s}^{l,0}_{x})  = \MyVec{s}^{l,0}_{x}.
     \end{aligned}
\end{equation}
%
$\MyVec{h}^{l,0}_x$ stores the sum of adjacent edge weights of all heads for $v_x$.
%
It is passed to the second sub-layer as ($\delta$ represents the activation function):
%
\begin{equation}
    \begin{aligned}
        \MyVec{m}^{l,1}_{y,x} &= \phi^{l,1}(\MyVec{h}^{l,0}_y, \MyVec{h}^{l,0}_x, \MyVec{e}_{y,x}, \hvec{y}, \hvec{x}) = \parallel_{k=1}^{K}\frac{\GATCalcWeight}{\MyVec{h}^{l,0}_x[k]}\hvec{y}[k], \\
        \MyVec{s}^{l,1}_x &= \mathlarger{\Sigma}_{v_y \in \mathcal{N}(v_x)} \MyVec{m}^{l,1}_{y,x}, \\
        \MyVec{h}^{l,1}_x &= \gamma^{l,1}(\MyVec{h}^{l,0}, \MyVec{s}^{l,1}_x) = \delta(\MyVec{s}^{l,1}_x).
    \end{aligned}
\end{equation}
%
In the GAT layer $l$, the graph neuron corresponding to $v_x$ first transforms the hidden vector of $v_x$ from the previous layer $\boldsymbol{h}^l_x$ with a weight matrix $\textcolor{blue}{\boldsymbol{W}^{l,k}}\boldsymbol{h}_x^l$.
calculates the weight (i.e., attention) of every adjacent edge dynamically with

@@

\begin{table}[h]

@@

\begin{table}

@@

                         & \boldsymbol{h}_x^{l+1} = \delta(\boldsymbol{s}_x^{l,1})

@@

                         & \boldsymbol{h}_x^{l+1} = \textcolor{blue}{\boldsymbol{W}^l_o} [\boldsymbol{h}_x^l \parallel (\boldsymbol{g}_{x}^l \odot \boldsymbol{h}_{x}^{l, 1})] + \textcolor{blue}{\boldsymbol{b}^l_o}

@@

\subsection{Classification of GNNs}

Since we focus on analyzing the performance bottleneck in training GNNs, we classify the typical GNNs from the view of time complexity.
We use $O_\phi$/$O_\Sigma$/$O_\gamma$ to denote the time complexity of the three functions in the message-passing model.
The time complexity of a GNN layer is made up of two parts: the edge calculation complexity $m * (O_\phi$ + $O_\Sigma)$ and the vertex calculation complexity $n * O_\gamma$.

In \tablename~\ref{tab:gnn_overview_edge} and \tablename~\ref{tab:gnn_overview_vertex}, we list the edge and vertex calculation complexity, respectively.
The time complexity of a graph neuron is affected by the dimensions of the input/output hidden vectors $d_{in}$ and $d_{out}$ and the dimensions of the model parameters (like the number of heads $K$ in GAT and the dimensions of the view vectors $d_a$/$d_v$/$d_m$ in GaAN).

We classify the typical GNNs into four quadrants based on their edge/vertex complexity as shown in \figurename~\ref{fig:gnn_complexity_quadrant}. We pick GCN, GGNN, GAT, and GaAN as the \emph{representative GNNs} of the four quadrants.

\begin{figure}
    \centering
    \includegraphics[width=0.7\columnwidth]{figs/illustration/GNN_complexity_quadrant.pdf}
    \caption{Complexity quadrants of typical GNNs. We compare the complexity according to the number of sub-layers, the Big-O notation and the number of parameters to train.}
    \label{fig:gnn_complexity_quadrant}
\end{figure}

\textbf{GCN} (low vertex \& edge calculation complexity): Graph convolution network (GCN \cite{kipf2017_gcn}) introduces the first-order approximation of the spectral-based graph convolutions.
%By combining the advantages of both the spectral-based and the spatial-based convolution, it can learn the representation of the local topological structure and vertex features efficiently.
It has only one parameter to learn at each layer, i.e. the weight matrix $\boldsymbol{W}^l$ in $\gamma$.
A GCN graph neuron can be expressed as $\boldsymbol{h}^{l+1}_i = \boldsymbol{W}^l\sum_{v_j \in \mathcal{N}(v_i)}{e_{j,i}\boldsymbol{h}^l_j}$, where $e_{j,i}$ is the normalized weight of the edge $(v_j, v_i)$.
%In the matrix view, $\boldsymbol{H}^{l+1} = (\boldsymbol{A}\boldsymbol{H}^l)\boldsymbol{W}^l$.
According to the associative law of the matrix multiplication, $\boldsymbol{h}^{l+1}_i = \sum_{v_j \in \mathcal{N}(v_i)}{e_{j,i}\boldsymbol{W}^l\boldsymbol{h}^l_j}$.
Since the dimension of $\boldsymbol{h}^{l+1}_i$ is usually smaller than $\boldsymbol{h}^l_i$ in practical GCNs, the implementation of GCN in PyTorch Geometric chooses to first conduct the vertex calculation $\hat{\boldsymbol{h}}^l_j = \boldsymbol{W}^l\boldsymbol{h}^l_j$ for each vertex $v_j$ and then conduct the edge calculation $\boldsymbol{h}^{l+1}_i=\sum_{v_j\in\mathcal{N}(v_i)}{e_{j,i}\hat{\boldsymbol{h}}^l_j}$.
As $\hat{\boldsymbol{h}}^l_j$ has the same dimension as $\boldsymbol{h}^{l+1}_i$, the implementation significantly reduces the  computation cost of the edge calculation.

\textbf{GGNN} (high vertex \& low edge calculation complexity):
GGNN \cite{li2015_ggnn} introduces the gated recurrent unit (GRU) into the graph neural networks.
The vertex updating function $\phi$ of GGNN is a modified GRU unit that has 12 model parameters to learn, having high computational complexity.
To lower the training cost, all GNN layers share the same group of parameters in GGNN.
GGNN further requires the dimension of $\boldsymbol{h}^{l+1}$ is equal to the dimension of $\boldsymbol{h}^l$.
Since the messaging function $\phi$ only uses the hidden feature vector $\boldsymbol{h}^l_j$ of the source vertex $v_j$ of an edge $(v_j, v_i)$. In the implementation of PyG, GGNN conducts the pre-processing vertex calculation $\hat{\boldsymbol{h}}^l_i=\boldsymbol{W}\boldsymbol{h}^l_i$ for every vertex $v_i$ before the message-passing.
The messaging function $\phi$ directly uses $\hat{\boldsymbol{h}}^l_j$ as the message vector for the edge $(v_j, v_i)$.
In this way, GGNN further reduces the time complexity of the edge calculation to $O(1)$ without increasing the time complexity of the vertex calculation.

\textbf{GAT} (low vertex \& high edge calculation complexity):
GAT \cite{huang2018_gat} introduces the attention and multi-head mechanism into the graph neural networks.
The $K$ heads generate $K$ independent views for an edge, where $K$ is a hyper-parameter.
The views of $K$ heads can be merged by concatenating or by averaging.
For concatenating, the dimension of the hidden feature vector of each head $d_{head}$ is $d_{out}/K$.
For averaging, $d_{head}$ is $d_{out}$, multiplying the complexity by $K$.
Each GAT layer consists of a preprocessing step and two sub-layers.
In the preprocessing step, GAT calculates the attention vectors $\hat{\boldsymbol{h}}^{l}_{i,(k)}$ for every vertex $v_i$ in every head $k$.
The first sub-layer uses the attention vectors to calculate the attention weights of every edge in every head $\boldsymbol{m}^{l,0}_{j,i,(k)}$.
After aggregation, the first sub-layer gets the summation of attention weights of every vertex in every head $\hat{\boldsymbol{a}}^{l,0}_{i,(k)}$.
The second sub-layer calculates the normalized attention weights $\alpha_{j, i, (k)}$ for every edge in every head and aggregates the hidden feature vectors with the normalized weights in the message-passing.
GAT uses the aggregated vector $\boldsymbol{s}^{l,1}_i$ of the second sub-layer as the output hidden vector directly.

\textbf{GaAN} (high vertex \& edge calculation complexity):
Based on the multi-head mechanism, GaAN \cite{zhang2018_gaan} introduces a convolutional subnetwork to control the weight of each head.
Each GaAN layer consists of four sub-layers. The first two sub-layers are similar to GAT, and the last two sub-layers build the convolutional subnetwork.
The first sub-layer aims to get the summation of attention weights of every vertex in each head $\boldsymbol{\hat{a}}^{l,0}_{j, i, (k)}$.
The second sub-layer aggregates the feature vectors with the normalized attention weights $\boldsymbol{\hat{a}}^{l,0}_{i,(k)}$.
The third and fourth sub-layers aggregate feature vectors with element-wise max and element-wise mean operators separately.
After the four sub-layers, the vertex updating funciton uses the aggregated vectors of last three sub-layers $\boldsymbol{s}^{l, 1}_i, \boldsymbol{s}^{l, 2}_i, \boldsymbol{s}^{l, 3}_i$ to generate the output hidden vector $\boldsymbol{h}^{l+1}_i$

>>>>

\begin{table}[h]
    \caption{Notations}
    \label{tab:notations}
    \centering
    \small
    \begin{tabular}{p{3em}lp{35em}}
        \toprule
       Category & Symbol & Meaning \\
        \midrule
        \multirow[c]{4}{3em}{Graph Structure}& $\mathcal{G}=(\mathcal{V}, \mathcal{E})$ & The simple undirected input graph with the vertex set $\mathcal{V}$ and the edge set $\mathcal{E}$. \\
        & $v_x$ & The $x$-th vertex of the input graph. \\
        & $e_{x,y}$ & The edge pointing from $v_x$ to $v_y$ of the input graph. \\
        & $\mathcal{N}(v_x)$ & The adjacency set of $v_x$ in the input graph. \\ \midrule
        \multirow[c]{6}{3em}{GNN Definition}& $L$ & The number of GNN layers. \\
        & $K$ & The number of heads in a GNN layer. \\
        & $\phi^l$ & The messaging function of the GNN layer $l$. \\
        & $\Sigma^l$ & The aggregation function of the GNN layer $l$. \\
        & $\gamma^l$ & The vertex updating function of the GNN layer $l$. \\ 
        & $\phi^{l,i}$ / $\Sigma^{l,i}$ / $\gamma^{l,i}$ & The messaging/aggregation/updating function of the $i$-th sub-layer of the GNN layer  $l$.\\
        & $\textcolor{blue}{\boldsymbol{W}^l, \boldsymbol{W}^{(k)}/\boldsymbol{b}, \boldsymbol{a}}$ & The matrices/vectors represented by the blue characters are the weight matrices/vectors that need to be learned in the GNN. \\  \midrule
        \multirow[c]{8}{3em}{Vectors}& $\boldsymbol{v}_x$ & The feature vector of the vertex $v_x$. \\
        & $\boldsymbol{e}_{x,y}$ & The feature vector of the edge $e_{x,y}$.  \\
        & $\boldsymbol{h}_x^{l}$ &  The {input} hidden vector of the graph neuron corresponding to $v_x$ in the GNN layer $l$. \\
        & $\boldsymbol{h}_x^{l+1}$ &  The {output} hidden vector of the graph neuron corresponding to $v_x$ in the GNN layer $l$.\\
        & $\boldsymbol{m}_{x,y}^l$ & The message vector of the edge $e_{x,y}$ outputted by $\phi^l$ of the GNN layer $l$. \\
        & $\boldsymbol{s}_{x}^l$ & The aggregated vector of the vertex $v_x$ outputted by $\Sigma^l$ of the GNN layer $l$. \\
        & $\boldsymbol{h}_{x}^{l,i}$ / $\boldsymbol{m}_{x,y}^{l,i}$ / $\boldsymbol{s}_{x}^{l,i}$ & The hidden/message/aggregated vector of the vertex $v_x$ outputted by $\gamma^{l,i}$/$\phi^{l,i}$/$\Sigma^{l,i}$ of the $i$-th sub-layer of the GNN layer $l$. \\
        & $d_{in}$, $d_{out}$ &  The dimensions of the input/output hidden vectors of a graph neuron. \\
        \bottomrule

@@

Graph neurons are building blocks of a GNN.
%
A GNN layer consists of $|\mathcal{V}|$ graph neurons.
%
Each vertex corresponds to a graph neuron.
%
A graph neuron is a small neural network.
%
For the graph neuron corresponding to $v_x$ at the layer $l$, it receives the hidden vector of itself $\boldsymbol{h}^l_x$ and the hidden vectors of its adjacent vertices $\boldsymbol{h}^l_y$ with $v_y \in \mathcal{N}(v_x)$ from the previous GNN layer\footnote{For the GNN layer 0, graph neurons receive input feature vectors, i.e., $\boldsymbol{h}^0_x=\boldsymbol{v}_x$}.
%
The graph neuron of $v_x$ aggregates the received hidden vectors, applies non-linear transformations, and outputs a new hidden vector $\boldsymbol{h}_x^{l+1}$.

@@

Different GNNs have different definitions of the three functions.
%
We regard $\phi$ and $\Sigma$ as \emph{edge calculation} functions, since they are conducted over every edge in $\mathcal{G}$.
%
We regard $\gamma$ as the \emph{vertex calculation} function, as it is conducted over every vertex in $\mathcal{G}$.
%
\tablename~\ref{tab:gnn_overview_edge} and \tablename~\ref{tab:gnn_overview_vertex} list the edge calculation functions and the vertex calculation functions of typical GNNs, respectively.
%
Some complex GNNs like GAT \cite{huang2018_gat} and GaAN  \cite{zhang2018_gaan} use more than one message passing phase in each GNN layer.
%
We regard every message passing phase in a GNN layer as a \emph{sub-layer}.
%
We will give out more details on sub-layers when we introduce GAT.

@@

\newcommand{\GATCalcWeight}{\exp(LeakyReLU(\Param{\MyVec{a}}^T[\hvec{y}[k] \parallel \hvec{x}[k]]))}

\begin{table}[H]

@@

\begin{table}[H]

@@

                         & \boldsymbol{h}_x^{l+1} = \MyVec{h}^{l,1}_x = \delta(\boldsymbol{s}_x^{l,1})

@@

                         & \boldsymbol{h}_x^{l+1} = \MyVec{h}^{l,3}_x = \textcolor{blue}{\boldsymbol{W}^l_o} [\boldsymbol{h}_x^l \parallel (\boldsymbol{g}_{x}^l \odot \boldsymbol{h}_{x}^{l, 1})] + \textcolor{blue}{\boldsymbol{b}^l_o}

@@

\subsection{Representative GNNs}

%Since we focus on analyzing the performance bottleneck in training GNNs, we classify the typical GNNs from the view of time complexity.
We use $O_\phi$/$O_\Sigma$/$O_\gamma$ to denote the time complexity of the three functions in the message-passing model.
The time complexity of a GNN layer is made up of two parts: the edge calculation complexity $O_\phi$ + $O_\Sigma$ and the vertex calculation complexity $O_\gamma$.
%
In \tablename~\ref{tab:gnn_overview_edge} and \tablename~\ref{tab:gnn_overview_vertex}, we list the edge and vertex calculation complexity of each GNN, respectively.
The time complexity of a graph neuron is affected by the dimensions of the input/output hidden vectors $d_{in}$ and $d_{out}$ and the dimensions of the model parameters (like the number of heads $K$ in GAT and the dimensions of the view vectors $d_a$/$d_v$/$d_m$ in GaAN).

Since we focus on analyzing the performance bottleneck in training GNNs, we classify the typical GNNs into four quadrants based on their edge/vertex complexity as shown in \figurename~\ref{fig:gnn_complexity_quadrant}. We pick GCN, GGNN, GAT, and GaAN as the \emph{representative} GNNs of the four quadrants.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\columnwidth]{figs/illustration/GNN_complexity_quadrant.pdf}
    \caption{Complexity quadrants of typical GNNs. We compare the complexity according to the number of sub-layers, the Big-O notation and the number of parameters to train.}
    \label{fig:gnn_complexity_quadrant}
\end{figure}

\subsubsection{GCN (Low Vertex \& Low Edge Complexity)}

Graph convolution network (GCN \cite{kipf2017_gcn}) introduces the first-order approximation of the spectral-based graph convolutions into graph neural networks.
%
It has only one parameter to learn at each layer, i.e. the weight matrix $\Param{\MyMat{W}^l}$ in the updating function $\gamma^l$.
%
A GCN graph neuron can be expressed as $\boldsymbol{h}^{l+1}_x = \Param{\MyMat{W}^l}\sum_{v_y \in \mathcal{N}(v_x)}{w_{y,x}\boldsymbol{h}^l_y}$, where $w_{y,x}$ is the normalized weight of the edge $e_{y,x}$.
%
According to the associative law of the matrix multiplication, $\boldsymbol{h}^{l+1}_x = \sum_{v_y \in \mathcal{N}(v_x)}{w_{y,x}\Param{\MyMat{W}^l}\boldsymbol{h}^l_y}$.
%
Since the dimension of $\boldsymbol{h}^{l+1}_x$ is usually smaller than $\boldsymbol{h}^l_x$ in practical GCNs, the implementation of GCN in PyG chooses to first conduct the vertex calculation $\hat{\boldsymbol{h}}^l_y = \Param{\MyMat{W}^l}\boldsymbol{h}^l_y$ for each vertex $v_y$ and then conduct the edge calculation $\boldsymbol{h}^{l+1}_x=\sum_{v_y\in\mathcal{N}(v_x)}{w_{y,x}\hat{\boldsymbol{h}}^l_y}$.
%
As $\hat{\boldsymbol{h}}^l_y$ has the same dimension as $\boldsymbol{h}^{l+1}_x$, the implementation significantly reduces the  computation cost of the edge calculation.

\subsubsection{GGNN (High Vertex \& Low Edge Complexity)}

GGNN \cite{li2015_ggnn} introduces the gated recurrent unit (GRU) into graph neural networks.
%
The updating function $\phi^l$ of GGNN is a modified GRU unit that has 12 model parameters to learn, having high computational complexity.
%
To lower the training cost, all GNN layers share the same group of parameters in GGNN.
%
GGNN further requires the dimension of $\boldsymbol{h}^{l+1}$ is equal to the dimension of $\boldsymbol{h}^l$.
%
Since the messaging function $\phi^l$ only uses the hidden vector $\boldsymbol{h}^l_y$ of the source vertex $v_y$ for an edge $e_{y,x}$, in the implementation of PyG, GGNN conducts the pre-processing vertex calculation $\hat{\boldsymbol{h}}^l_x = \Param {\MyMat{W}^l}\boldsymbol{h}^l_x$ for every vertex $v_x$ before the message passing.
%
The messaging function $\phi^l$ directly uses $\hat{\boldsymbol{h}}^l_y$ as the message vector for every edge $e_{y, x}$.
%
In this way, GGNN further reduces the time complexity of the edge calculation to $O(1)$ without increasing the time complexity of the vertex calculation.

\subsubsection{GAT (Low Vertex \& High Edge Complexity)}

GAT \cite{huang2018_gat} introduces the multi-head attention mechanism into graph neural networks.
%
Each GAT layer has $K$ heads that generate $K$ independent views for an edge, where $K$ is a hyper-parameter.
%
The views of $K$ heads can be merged by concatenating or by averaging.
%
For concatenating, the dimension of the hidden vector of each head $d_{head}$ is $d_{out}/K$.
%
For averaging, $d_{head}$ is $d_{out}$.

Each GAT layer consists of a vertex pre-processing phase and two sub-layers.

The vertex pre-processing phase calculates the attention vector $\hat{\boldsymbol{h}}^{l}_{x}$ for every vertex $v_x$ by $\hvec{x} = \parallel_{k=1}^K \Param{\MyMat{W}^l_{(k)}}\MyVec{h}^l_x$. We denote the attention sub-vector generated by the $k$-th head as $\hvec{x}[k]=\Param{\MyMat{W}^l_{(k)}}\MyVec{h}^l_x$.

The two sub-layers also follow the message-passing model.
The first sub-layer (defined in Equation~\ref{eq:GAT-sub-layer-1}) uses the attention vectors to emit the attention weight vector for every edge $\boldsymbol{m}^{l,0}_{y,x}$ and aggregates the attention weight vectors for every vertex $v_x$ to get the weight sum vector $\MyVec{h}^{l,0}_x$.
%
\begin{equation}
    \label{eq:GAT-sub-layer-1}
    \begin{aligned}
        \MyVec{m}^{l,0}_{y,x} & = \phi^{l,0}(\MyVec{h}^l_y, \MyVec{h}^l_x, \MyVec{e}_{y,x}, \hvec{y}, \hvec{x}) = \parallel_{k=1}^{K}\GATCalcWeight, \\
        \MyVec{s}^{l,0}_{x} &= \mathlarger{\Sigma}_{v_y \in \mathcal{N}(v_x)}{\MyVec{m}^{l,0}_{y,x}}, \\
        \MyVec{h}^{l,0}_{x} &= \gamma^{l,0}(\MyVec{h}^l_x, \MyVec{s}^{l,0}_{x})  = \MyVec{s}^{l,0}_{x}.
    \end{aligned}
\end{equation}
%
The second sub-layer (defined in Equation~\ref{eq:GAT-sub-layer-2}) uses the weight sum vectors to normalize the attention weights for every edge and aggregates the attention vectors $\boldsymbol{\hat{h}}^l_y$ with the normalized weights.
%
The aggregated attention vectors $\MyVec{s}^{l,1}_x$ are transformed by an activation function $\delta$ and outputted as the hidden vectors of the current layer $\MyVec{h}^{l+1}_x$.
%
\begin{equation}
        \label{eq:GAT-sub-layer-2}
    \begin{aligned}
        \MyVec{m}^{l,1}_{y,x} &= \phi^{l,1}(\MyVec{h}^{l,0}_y, \MyVec{h}^{l,0}_x, \MyVec{e}_{y,x}, \hvec{y}, \hvec{x}) = \parallel_{k=1}^{K}\frac{\GATCalcWeight}{\MyVec{h}^{l,0}_x[k]}\hvec{y}[k], \\
        \MyVec{s}^{l,1}_x &= \mathlarger{\Sigma}_{v_y \in \mathcal{N}(v_x)} \MyVec{m}^{l,1}_{y,x}, \\
        \MyVec{h}^{l+1}_x = \MyVec{h}^{l,1}_x &= \gamma^{l,1}(\MyVec{h}^{l,0}, \MyVec{s}^{l,1}_x) = \delta(\MyVec{s}^{l,1}_x).
    \end{aligned}
\end{equation}

\subsubsection{GaAN (High Vertex \& High Edge Complexity)}

Based on the multi-head mechanism, GaAN \cite{zhang2018_gaan} introduces a convolutional subnetwork to control the weight of each head.
%
Each GaAN layer consists of four sub-layers (as defined in Table~\ref{tab:gnn_overview_edge} and Table~\ref{tab:gnn_overview_vertex}).
%
The first two sub-layers are similar to GAT, and the last two sub-layers build the convolutional subnetwork.
%
The first sub-layer aims to get the sum of attention weights of every vertex in all heads $\MyVec{h}^{l,0}_x$.
The second sub-layer aggregates the hidden vectors from the previous GaAN layer with the normalized attention weights for all heads $\alpha_{(k)}$.
The third and fourth sub-layers aggregate the hidden vectors from the GaAN previous layer with the element-wise max and mean operators separately.
The vertex updating funciton $\gamma^{l,4}$ of the fourth sub-layer uses the hidden vectors of the last two sub-layers $\MyVec{h}^{l, 1}_x, \MyVec{h}^{l, 2}_x$ to generate the output hidden vector $\MyVec{h}^{l+1}_x$.
